Sperm should evolve to make female meiosis fair.

### 2014-07-18 writeup_and_notes/GeneticsResub/male_dependent_drive.tex
Changes to intro

{\bf Abstract:}
Genomic conflicts arise when an allele gains an evolutionary advantage at a cost to organismal fitness. 
O\"{o}genesis is inherently susceptible to such conflicts
because alleles compete to be the  product of female meiosis transmitted to the egg. 
Alleles that distort meiosis in their favor (i.e. meiotic drivers) often decrease organismal fitness,
	and therefore indirectly favor the evolution of mechanisms to suppress meiotic drive. 
In this light, many facets of  o\"{o}genesis and gametogenesis have been 
	interpreted as mechanisms of protection against genomic outlaws. 
\yb{The fact that female meiosis often left uncompleted until after
      fertilization in many animals  appears to run counter to this interpretation, 
      as this delay provides an opportunity for sperm alleles to
     meddle with its outcome and help like-alleles drive in heterozygous females. } 
%Why then is female meiosis often left uncompleted until after
%fertilization in many animals -- 
%	potentially providing an opportunity for sperm alleles to
%     meddle with its outcome and help like-alleles drive in heterozygous females? 
The population genetic theory presented herein suggests that
	sperm nearly always evolve to increase the fairness of female
        meiosis in the face of genomic conflicts. 
These results are consistent with current knowledge of sperm-dependent meiotic drivers 
(loci whose distortion of female meiosis depends on sperm genotype). 
\yb{Rather than providing an opportunity for sperm collaboration in female meiotic drive,
        the requirement of fertilization for the completion of female meiosis
        likely indirectly provides protection to females against meiotic drivers as it provides an opportunity for sperm to suppress female meiotic drive.}
%	and suggest that the requirement of fertilization for the
%%        completion of female meiosis
%        potentially represents a mechanism employed by females 
%	 to ensure a fair meiosis.
\newpage

@@

It is therefore somewhat surprising that despite the intense evolutionary pressure on female meiosis to prevent meiotic drive, 
it is potentially open to sabotage by a virtual stranger -- a haploid sperm genome.
That is, in many animal species, female meiosis is completed only after fertilization \citep{Masui_book}, 
	creating ample opportunity for interaction between the sperm and
	female meiotic machinery (note that the alternation of generations
	likely precludes this interaction in plants). 
\gc{It is becoming
        increasingly clear that sperm bring a wide variety of RNA and proteins into the
        egg REF. Some of these have known functions, for example, in most
        animal species sperm-- not eggs-- are
        responsible for the transmission of the centrosome, a vitial
        component of the mitotic machinary for the zygote. 
Detailed functional studies and paternal effect mutations in model
systems further highlight
 that sperm-transmitted products have a wide-range of functions in
 the completion of syngamy, zygotic development, and the resumption of
 female meiosis \cite{Miller:01, McNally:05,REF}. 
However,
        the function of many of the products the sperm brings into the
        egg is completely unknown and 
      these products vary widely over species. 
It seems quite plausible that sperm based products and
henece alleles could potentially have functional effects on
various aspects of female meiosis that occur after fertilization. } 
Therefore, a `green-bearded' \citep{Gardner2010} sperm-acting allele that recognizes and facilitates the meiotic drive of a genetically equivalent allele in heterozygous females %female meiosis %(a  heteromorphic dyad), 
     could presumably rapidly spread through a population. 
At first sight, female meiosis appears primed for \gc{conflict caused by
the such selfish systems.} Here we ask if sperm do indeed evolve to collaborate with female drivers to exploit this apparent weakness in the defense against meiotic drive. \newline 
%Why then is the requirement of fertilization to complete female meiosis so ubiquitous? 
\gc{THERE may be too much bio. detail in previous paragraph}

It is not the case that sperm is mechanistically incapable of influencing female meiosis.
Both mechanistic and genetic evidence suggests that sperm can influence the outcome of female meiosis.
An example of mechanistic evidence for sperm's influence on female meiosis comes from \emph{C. elegans}, where premature deployment of the aster -- a vital 	component of mitotic machinery provided by the sperm, 
	disrupts MII meiotic segregation in the egg, leading to a triploid zygote \citep{McNally2012}.
There are current only two polymorphic,characterized females meiotic drive systems in
animals-- and both of them, the \emph{In} and \emph{Om}  drive systems in mice-- suggests that the 
	 transmission of alternate alleles in heterozygous females can depend on sperm haplotype.  
 While ruling out the alternative hypothesis of early selection on zygotes in these cases is challenging  \citep[see pages
	52-54 in][ for comment]{Burt2006}, it appears that the extent to which \emph{In} and \emph{Om} distort the second female meiotic division partially depends on the 	
	genotype of the fertilizing sperm
        \citep{Agulnik1993,Wu2005}. 
The fact that both of polymorphic, putative female meiotic drive systems identified
to date in animals (\emph{In} and \emph{Om} in mice) seem to show an
effect of sperm genotype suggests that if female meiotic drive is
common the role of sperm will be important as well.

>>>>

{\bf Abstract:}
Genomic conflicts arise when an allele gains an evolutionary advantage at a cost to organismal fitness. 
O\"{o}genesis is inherently susceptible to such conflicts
because alleles compete to be the  product of female meiosis transmitted to the egg. 
Alleles that distort meiosis in their favor (i.e. meiotic drivers) often decrease organismal fitness,
	and therefore indirectly favor the evolution of mechanisms to suppress meiotic drive. 
In this light, many facets of  o\"{o}genesis and gametogenesis have been 
	interpreted as mechanisms of protection against genomic outlaws. 
The fact that female meiosis often left uncompleted until after
      fertilization in many animals appears to run counter to this interpretation, 
      as this delay provides an opportunity for sperm alleles to
     meddle with its outcome and help like-alleles drive in heterozygous females. 
%Why then is female meiosis often left uncompleted until after
%fertilization in many animals -- 
%	potentially providing an opportunity for sperm alleles to
%     meddle with its outcome and help like-alleles drive in heterozygous females? 
The population genetic theory presented herein suggests that
	\gc{in fact} sperm nearly always evolve to increase the fairness of female
        meiosis in the face of genomic conflicts. 
These results are consistent with current knowledge of sperm-dependence
at the \gc{for the well characterized female meiotic drivers in animals to date.} 
Rather than providing an opportunity for sperm collaboration in female meiotic drive,
        the requirement of fertilization for the completion of female meiosis
        likely indirectly provides protection to females against
        meiotic drivers as it provides an opportunity for sperm to
        suppress female meiotic drive. }
%	and suggest that the requirement of fertilization for the
%%        completion of female meiosis
%        potentially represents a mechanism employed by females 
%	 to ensure a fair meiosis.
\newpage

@@

It is therefore somewhat surprising that despite the intense evolutionary pressure on female meiosis to prevent meiotic drive, 
it is potentially open to sabotage by a virtual stranger -- a haploid sperm genome.
That is, in many animal species, female meiosis is completed only after fertilization \citep{Masui_book}, 
	creating ample opportunity for interaction between the sperm and
	female meiotic machinery (note that the alternation of generations
	likely precludes this interaction in plants). 
\gc{It is becoming
        increasingly clear that sperm bring a wide variety of RNA and proteins into the
        egg REF. Some of these have known functions, for example, in most
        animal species sperm-- not eggs-- are
        responsible for the transmission of the centriole, a vitial
        component of the mitotic machinary for the zygote. 
Detailed functional studies and paternal effect mutations in model
systems further highlight
 that sperm-transmitted products have a wide-range of functions in
 the completion of syngamy, zygotic development, and the resumption of
 female meiosis \cite{Miller:01, McNally:05,REF}. 
However,
        the function of many of the products the sperm brings into the
        egg is completely unknown and 
      these products vary widely over species \citep{Karrbook:09}. 
It seems quite plausible that sperm-based products and
hence alleles could potentially have functional effects on
various aspects of female meiosis that occur after fertilization. } 
Therefore, a `green-bearded' \citep{Gardner2010} sperm-acting allele that recognizes and facilitates the meiotic drive of a genetically equivalent allele in heterozygous females %female meiosis %(a  heteromorphic dyad), 
     could presumably rapidly spread through a population. 
At first sight, female meiosis appears primed for \gc{conflict caused by
the such selfish systems.} Here we ask if sperm do indeed evolve to collaborate with female drivers to exploit this apparent weakness in the defense against meiotic drive. \newline 
%Why then is the requirement of fertilization to complete female meiosis so ubiquitous? 
\gc{THERE may be too much bio. detail in previous paragraph, I can
  add more refs. once we've decided on content}

It is certainly not the case that sperm is mechanistically incapable of influencing female meiosis.
Both mechanistic and genetic evidence suggests that sperm can influence the outcome of female meiosis.
An example of mechanistic evidence for sperm's influence on female meiosis comes from \emph{C. elegans}, where premature deployment of the aster -- a vital component of mitotic machinery provided by the sperm, 
	disrupts MII meiotic segregation in the egg, leading to a triploid zygote \citep{McNally2012}.
There are current only two polymorphic,characterized females meiotic drive systems in
animals-- and both of them, the \emph{In} and \emph{Om}  drive systems in mice-- suggests that the 
	 transmission of alternate alleles in heterozygous females can depend on sperm haplotype.  
 While ruling out the alternative hypothesis of early selection on zygotes in these cases is challenging  \citep[see pages
	52-54 in][ for comment]{Burt2006}, it appears that the extent to which \emph{In} and \emph{Om} distort the second female meiotic division partially depends on the 	
	genotype of the fertilizing sperm
        \citep{Agulnik1993,Wu2005}. 
The fact that both of polymorphic, putative female meiotic drive systems identified
to date in animals (\emph{In} and \emph{Om} in mice) seem to show an
effect of sperm genotype suggests that if female meiotic drive is
common the role of sperm will be important as well.
